---
title: "TTO"
author: "Rohan Patel"
date: "2023-03-01"
output: html_document
---

A Statistical Analysis on Three-True Outcome Baseball (Plate Appearances ending in a walk, strikeout, or home run), as well as its merits and sustainability. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggpubr)
```

Looking at how TTO% (percent of PA that end in a TTO) has grown historically (Data from Fangraphs)
```{r}
historical <- read_csv("TTOHistorical.csv")


colnames(historical)[2]="TTO"

df<-data.frame(historical)
df$new_TTO<-gsub("%","",df$TTO)
historicalnew <- df

historicalnew$new_TTO <- as.numeric(historicalnew$new_TTO)/100

historicalnew


historical1 <- filter(historical,Season >= 1980)


ggplot(historicalnew, aes(x=Season,y=new_TTO))+
  geom_line()+
  labs(x="Season",y="Three True Outcome Percentage",title="League-wide Three True Outcome Percentage in MLB from 1930-2022")
```

2022 TTO Data (from Fangraphs):
- Filter for hitters with a TTO% greater than 40 in 2023 (large number)
```{r}
TTO2022 <- read_csv("TTO2022.csv")

colnames(TTO2022)[2]="Team"
colnames(TTO2022)[3]="TTO"


TTO1 <- filter(TTO2022,TTO >= 40,)
arrange(TTO1, desc(TTO))

```

Graphing TTO% vs wRC+ (Is relying on TTO-baseball good for offensive production?)
```{r}
colnames(TTO2022)[4] = "wRCplus"


df<-data.frame(TTO2022)
df$new_TTO<-gsub("%","",df$TTO)
TTO2022new <- df

TTO2022new$new_TTO <- as.numeric(TTO2022new$new_TTO)/100

TTO2022new



plot <- ggscatter(TTO2022new,x="new_TTO",y="wRCplus",add="reg.line",title="2022 wRC+ vs 2022 TTO% (Qualified)",xlab = "TTO%",ylab = "wRC+")

plot + stat_cor(method="pearson") 
  
  
```


TTO% isn't correlated with offensive success, so the next step is to try and answer why. The answer may be that reliance on TTO baseball is less sustainable because HR/FB rates fluctuate due top extraneous factors. 

Below we are creating a two-year data set (from Fangraphs) which allows us to compare wRC+ in 2021 and 2022 for hitters (to set a baseline for sustainability). We limited the analysis to hitters with at least 300 PAs in each year, so some names had to be removed. 

```{r}
TTO3 <- read.csv("2yearwRCTTO.csv")

df<-data.frame(TTO3)
df$new_TTO<-gsub("%","",df$TTO)
TTO4 <- df


TTO4$new_TTO <- as.numeric(TTO4$new_TTO)/100

TTO2021 <- filter(TTO4, Season==2021, new_TTO >= .40)

TTO2021

vec_names <- TTO2021$Name

vec_names_exclude <- c("Brandon Lowe","Chris Taylor","Bryce Harper","Adam Duvall","Fernando Tatis Jr.","Giancarlo Stanton","Joey Gallo","Miguel Sano","Joey Votto","Robbie Grossman","Ryan Mountcastle","Tyler O'Neill","Yoan Moncada")

vec_names1 <- setdiff(vec_names,vec_names_exclude)

vec_names1

TTO5 <- filter(TTO4, Name %in% vec_names1) 
TTO6 <- arrange(TTO5,Name)

TTO6

colnames(TTO6)[5] = "TTO"

TTO6
write_csv(TTO6,"TT06.csv")
```

```{r}
TTO6_sorted <- read_csv("TTO6sorted.csv")


colnames(TTO6_sorted)[5] = "wRCplus2"
colnames(TTO6_sorted)[3] = "wRCplus1"

TTO6_sorted

```

The next couple chunks help to create smaller datasets (one for >40% TTO), as well as graph the baseline correlation for wRC+

```{r}
TTO7 <- read_csv("TTOexpanded.csv")

colnames(TTO7)[5]="TTO"
colnames(TTO7)[4]="wRCplus"

df<-data.frame(TTO7)
df$new_TTO<-gsub("%","",df$TTO)
TTO7 <- df

TTO7$new_TTO <- as.numeric(TTO7$new_TTO)*100



TTO8 <- filter(TTO7, Season=="2021",new_TTO >= .40)

vec_names1 <- TTO8$Name

vec_names1

TTO9 <- filter(TTO7, Name %in% vec_names1)

TTO10 <- arrange(TTO9,Name)

n_occur <- data.frame(table(TTO10$Name))

n_occur1 <- n_occur[n_occur$Freq > 1,]


vec_names2 <- c(n_occur1$Var1)


TTO11 <- filter(TTO10,Name %in% vec_names2)

TTO11

write_csv(TTO11,"TTO11.csv")
```

```{r}
TTO7

n_occur2 <- data.frame(table(TTO7$Name))

n_occur3 <- n_occur2[n_occur2$Freq > 1,]

vec_names3 <- n_occur3$Var1

TTO13 <- filter(TTO7,Name %in% vec_names3)

write_csv(TTO13,"TTO13.csv")

```

Graph of baseline wRC+ correlation from 2021-2022 (so we can compare the results for the various TTO% groups)

```{r}
TTO14 <- read_csv("TTO13sorted.csv")

plot <- ggscatter(TTO14,x="wRCplus2021",y="wRCplus2022",add="reg.line",title="2022 wRC+ vs 2021 wRC+ (min 300 PA))",xlab = "2021 wRC+",ylab="2022 wRC+")

plot + stat_cor(method="pearson") 

```

Graph and correlation for the group with a high TTO% (>40%):

```{r}


TTO12 <- read_csv("TTO11sorted.csv")

ggplot(TTO12,aes(x=wRC2021,y=wRC2022))+
  geom_point()

cor.test(TTO12$wRC2022,TTO12$wRC2021,method="pearson")


plot <- ggscatter(TTO12,x="wRC2021",y="wRC2022",add="reg.line",title="2022 wRC+ vs 2021 wRC+ for Hitters with TTO% above 40 (min 300 PA)",xlab = "2021 wRC+",ylab="2022 wRC+") 

plot + stat_cor(method="pearson") 


```

Clearly, the correlation for this group is much smaller than the baseline, which suggests that wRC+ is significantly less sticky for the group of hitters with a hogh TTO%. The explanation may lie in the year-to-year fluctuations in HR/FB rates. 

```{r}
HRFB <- read_csv("HR.csv")

colnames(HRFB)[22] = "HRFB"

df<-data.frame(HRFB)
df$HRFB<-gsub("%","",df$HRFB)
HRFB <- df



ggplot(HRFB,aes(x=Season,y=HRFB,"HR/FB Rate by Year (2002-2022)"))+
  geom_line()+
  xlab("Year")+
  ylab("HR/FB Rate")
```


Creating a smaller dataset and graph for the group with a low TTo% (<30%)
```{r}
TTO15 <- read_csv("TTO132.csv")

df<-data.frame(TTO15)
df$new_TTO2022<-gsub("%","",df$TTO2022)
TTO15 <- df

df<-data.frame(TTO15)
df$new_TTO2021<-gsub("%","",df$TTO2021)
TTO15 <- df

df<-data.frame(TTO15)
df$new_TTOdiff<-gsub("%","",df$TTOdiff)
TTO15 <- df



TTO15

TTO16 <- filter(TTO15,new_TTO2021 <= 30)

TTO16

plot <- ggscatter(TTO16,x="wRCplus2021",y="wRCplus2022",add="reg.line",title="2022 wRC+ vs 2021 wRC+ for Hitters with TTO% below 30 (min 300 PA)",xlab = "2021 wRC+",ylab="2022 wRC+")

plot + stat_cor(method="pearson") 

```

wRC+ is also significantly less sticky than the baseline for the group with a low TTO%. Generally, hitters with a low TTO% rely more on balls in play which are suceptible to significant variance due toq extraneous factors. While league-wide BABIP rarely changes, individual hitters' BABIPs often fluctuate affecting wRC+


Smaller Dataset for the group with the "middle" TTO% (30-40%).

```{r}
TTO17 <- filter(TTO15, 30 <= new_TTO2021 & new_TTO2021 <= 40 )

TTO17

plot <- ggscatter(TTO17,x="wRCplus2021",y="wRCplus2022",add="reg.line",title="2022 wRC+ vs 2021 wRC+ for Hitters with 30-40 TTO% (min 300 PA)",xlab = "2021 wRC+",ylab="2022 wRC+")

plot + stat_cor(method="pearson") 
```

Unlike the other two groups, the correlation is significantly stronger than the baseline. This suggests that offensive production is more sustainable when TTO% is between these range. Hitters in this range rely on both TTO and Balls in Play and creates a sustainable balance. 


Below are some player examples that I touched on in my article on Medium: 

https://medium.com/@xRPx7/feast-or-famine-analyzing-the-sustainability-of-three-true-outcome-baseball-f2f86df903b2


```{r}
Yuli <- read_csv("Yuli.csv")

colnames(Yuli)[5] = "wRCplus"

ggplot(Yuli, aes(x=Season,y=BABIP))+
  geom_line()
ggplot(Yuli, aes(x=Season,y=wRCplus))+
  geom_line()

arrange(Yuli,Season)
```

```{r}
Gallo <- read_csv("Gallo.csv")

colnames(Gallo)[4] = "wRCplus"
colnames(Gallo)[5] = "TTO"

Gallo1<- filter(Gallo, Season >= 2017)

as.numeric(Gallo1$TTO)

is.numeric(Gallo1$TTO)




ggplot(Gallo1, aes(x=Season,y=TTO))+
  geom_line()
ggplot(Gallo1, aes(x=Season,y=wRCplus))+
  geom_bar(stat="identity") 

arrange(Gallo1,Season)

```

To conclude, these findings help us find which offensive approach is more sustainable, not necessarily what approach is more sucessful. Ultimately, the offensive approach a hitter takes depends on their profile. For example, Yankees Superstar Aaron Judge routinely runs high TTO%, but consistently has great offensive output. Judge hits a lot of extremely long home runs, which makes him less suceptible to HR/FB variance. 
